Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-10 paper/ms.tex
update ms

%%%%%%%%%
% STARS %
%%%%%%%%%
\input{vals_manuscript}

% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\approx$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\approx$6\%}}

%%%%%%%%%%%
% PLANETS %
%%%%%%%%%%%
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

@@

%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
%\title{Ages of the Kepler Stars and Planets}
%\title{A Dearth of Young Stars and Planets in the Kepler Field}
\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}

@@

\keywords{Stellar ages (1581), Planet hosting stars (1242), Field
stars (2103), Exoplanet evolution (491), Milky Way evolution (1052),}

@@

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for transiting planet host star ?
Second, where are the missing young planets?  Briefly, we will find
that uniform is incorrect by a factor of a few, and that at least one
bias against the discovery of young transiting planets is in the
measurement of their ages.

Given that Kepler's main mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged isochronal ages from
an earlier analogous study \citep{Fulton_2018}, and also
used a relative gyrochrone age-ranking approach.
%that sorted stars
%based on whether they rotated faster or slower than a gyrochrone age
%of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
These studies were conducted in the broader context of investigating
trends in exoplanet demographics, which also depend on stellar
metallicity and mass \citep[e.g.][]{Petigura_2018}.  These studies
offered limited sensitivity to the ages of $\lesssim$1\,Gyr planets,
due to the nature of isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
have also been reported \citep{2021AJ....161..189L}.
There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the new calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
These improvements form the basis for why we believe the present
analysis to be necessary.
%Today, FGK star rotation-based ages are
%empirically tied to the cluster age scale, and
%are limited in precision by the intrisic scatter in the $P_{\rm rot}$
%sequences, and in accuracy by systematic uncertainties in the evolving
%spin-down rates, and in the cluster ages themselves.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Rotation periods for Kepler target stars (left) and known
			planet hosts (right)}.  
		The gray lines are ``mean fits'' 
		to the slow rotation sequences of open
		clusters.  The stellar sample (left) includes only apparently single stars
		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
		and temperatures of 3800--6200\,K.
		The color map has a linear stretch.  %between one and ten stars per bin.
		The planet-hosts (right) require
		the same stellar cuts, and include only the confirmed and candidate planets
		described in Section~\ref{subsec:plflags}.
		An order of magnitude estimate for the number of young planets discovered
		by Kepler follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}. 
%A relevant flag for
%stars missing \citetalias{McQuillan_2014} periods is included in
%Table~\ref{tab:stars}.

@@

We considered planets in the NEA cumulative KOI
table, which included the best knowledge available on any given planet
candidate as of 6 June 2023, while also incorporating human-based
vetting. %\footnote{See
%\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
These planets represent a superset of those in the fully automated
Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
future work for planet occurrence rate calculations.  This version of
the cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

% optional todo's:
%TODO: plot uncertainties in teff vs teff...
%TODO: plot uncertainties in logg vs logg...
%TODO: add that plot showing differences btwn the two teff scales?

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all
\nuniqstarsantosrotteffcut\ stars with reported rotation periods and
effective temperatures of 3800--6200\,K.  To flag stars for
which we suspect these ages may not in fact be valid, we built a set
of quality flags, which we condensed into a single binary number,
$Q_{\rm star}$.  When and how the flags represented in this should be
applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they could all be applied.  If
the goal is to construct a complete sample, then consider the examples
of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
($\approx$625\,Myr).  The former has a high RUWE due to a resolved
binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
\citep{2014ApJ...783...53M}.  We leave selection for or against such
cases as a decision to the user.

@@

{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3 renormalized
unit weight error
(RUWE) as a function of other stellar parameters, %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
and flag stars with RUWE$>$1.4 as possible binaries.  Such astrometric
outliers can be either bona fide astrometric binaries, or more often
are marginally resolved point sources for which a single-source PSF
model provides a poor fit.

@@

{\it Candidate pulsators and close-in binaries (bit
10)}---\citeauthor{Santos_2021} included in a flag for candidate
``classical pulsators'' (e.g.\ Cepheids) and
close-in binaries.  Visual inspection 
% of the rotation periods as a function of temperature 
shows that although this flag selects many bona fide objects in these
classes, it also selects most $P_{\rm
rot}$$\lesssim$5\,day bona fide young planets.  
These objects are neither classical pulsators nor close
binaries.  While we propagated this flag into our set of quality
flags, we therefore generally ignore it.

@@

%Certain quality flags described in Section~\ref{subsec:flags} are also
%relevant for lithium ages, specifically those associated with
%crowding and binarity.
%For instance, rapid rotators are more common in binaries.
%and they also
%generally have higher lithium abundances than slow-rotating stars of the
%same mass (CITE).
%Stars for which rotation-based age dating is dubious heavily overlap
%with those for which lithium-based ages are dubious.

@@

%The new young planets
%include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
%(Kepler-1529b $\approx$100\,Myr,
%-1521b $\approx$195\,Myr,
%-1467b $\approx$235\,Myr,
%-955b $\approx$400\,Myr, and
%-523b $\approx$440\,Myr),
%a single Neptune-sized planet
%(Kepler-971b $\approx$335\,Myr),
%five 1.1-1.8\,$R_\oplus$ super-Earths
%(Kepler-1930b $\approx$150\,Myr,
%-1313b $\approx$185\,Myr,
%-1565b $\approx$205\,Myr,
%-1532b $\approx$400\,Myr,
%-1481b $\approx$400\,Myr),
%and three Earth or sub-Earth-sized planets
%(Kepler-1933b $\approx$300\,Myr,
%-1561b $\approx$420\,Myr,
%-1482b $\approx$490\,Myr).

@@

%Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
%1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
%$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
%though at half the relative uncertainty.

%A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
%planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
%find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.

@@

%\subsubsection{Need spectra}
%Kepler-1938...
%Kepler-762... (a HJ!)
%Kepler-1320

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates of main sequence stars.
Binarity is one example: even intermediate-separation ($\sim$10\,AU) binaries are biased
toward rapid rotation
\citep[e.g.][and many studies thereafter]{Meibom_2007}. %,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
In this work, we were interested in not only the overall
age distribution of the Kepler field, but also in the reliability of individual ages for
young planet hosting stars.
The quality flagging scheme in Table~\ref{tab:planets} provides one path toward this end.
The other method is to consider the region in which we do have
overlapping sensitivity between lithium and rotation -- 
G and K dwarfs with $t$$\lesssim$1\,Gyr.  
Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in agreement
(with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a 
precise lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

In this section we discuss fifteen systems with confirmed planets that have nominally
discrepant rotation and lithium-based ages.
We find that only one of them, Kepler-786, is interesting.

@@

%
%Our quality flags noted that this star has
%a low surface gravity, a crowded field, high RUWE, and an ambiguous Kepler-to-Gaia crossmatch.

@@

%  Kepler-63b is interesting in that it is {\it missing} from our
%  catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
%  a polar orbit based on both the Rossiter-McLaughlin effect and a
%  starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
%  eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
%  a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
%  qualitatively consistent with the estimates from
%  \citet{2013ApJ...775...54S}, though with larger uncertainties because
%  our age accounts for the intrinsic scatter in the open cluster
%  calibration sequences \citep{Bouma_2023}.  
%  While the usual catalogs (\citetalias{Santos_2019},
%  \citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
%  \citetalias{Mazeh_2015}) do not report this star's rotation period.
%  \citetalias{Santos_2021} flagged the system as ``potentially'' having
%  rotation-based modulation.

>>>>

%%%%%%%%%%
% VALUES %
%%%%%%%%%%
\input{vals_manuscript}

% number of stars monitored by Kepler (quoting Santos19/21)
\newcommand{\nkeplerstars}{$\approx$160{,}000}
% fraction of stars with rotation periods, with teff/logg in Berger+20
\newcommand{\fracstarswithprotwithbtwenty}{{$\approx$94\%}}
% complement of above
\newcommand{\fracstarswithprotwithoutbtwenty}{{$\approx$6\%}}
% fraction of KOIs that are not FPs with B+20 parameters
\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}

@@

\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}

@@

\keywords{Stellar ages (1581), Planet hosting stars (1242), Field
stars (2103), Exoplanet evolution (491), Milky Way evolution (1052)}

@@

This study aims to resolve two main questions.  First, how wrong is it
to assume a uniform age distribution for transiting planet host stars?
Second, where are the missing young planets?  Briefly, we will find
that uniform is wrong by a factor of a few, and that at least one bias
against the discovery of young planets is in the measurement of their
ages.

Given that Kepler's main mission ended in 2013, one might be forgiven
for assuming that the ages of the Kepler planets are already known.
Of course, they have been studied: \citet{Berger_2020b_rpage} and
\citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged isochronal ages from an
earlier analogous study \citep{Fulton_2018}, and also used a relative
gyrochrone age-ranking approach.  These studies were conducted in the
broader context of investigating trends in exoplanet demographics,
which also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of $\lesssim$1\,Gyr planets, due to the nature
of isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic''
ages have also been reported \citep{2021AJ....161..189L}.  There have
been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our
    vetting of false positives within it;
  %
  \item Available measurements of stellar rotation periods in open
    clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time \citep{Bouma_2023}.
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  For late G dwarfs and
    K dwarfs, the new calibration differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
These improvements form the basis for why we believe the present
analysis to be necessary.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  
    The gray lines are ``mean fits'' to the slow rotation sequences of
    open clusters.  The stellar sample (left) includes only apparently
    single stars near the main sequence with $\log g$$>$4.2,
    RUWE$<$1.4, and temperatures of 3800--6200\,K.  The color map has
    a linear stretch.  The planet-hosts (right) require the same
    stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.  An order of magnitude
    estimate for the number of young planets discovered by Kepler
    follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}.

@@

We considered planets in the NEA cumulative KOI table, which included
the best knowledge available on any given planet candidate as of 6
June 2023, while also incorporating human-based vetting.  These
planets represent a superset of those in the fully automated Q1-Q17
DR25 KOI Table \citep{Thompson_2018}, which could be adopted in future
work for planet occurrence rate calculations.  This version of the
cumulative KOI table included \nkoisnofp\ objects that are either
``confirmed'' or ``candidate'' planets, after excluding known false
positives.

@@

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
stars with reported rotation periods and effective temperatures of
3800--6200\,K.  To flag stars for which we suspect these ages may not
in fact be valid, we built a set of quality flags, which we condensed
into a single binary number, $Q_{\rm star}$.  When and how the flags
represented in this should be applied depends on the question being
asked.  If the goal is to construct a false-positive free sample, they
could all be applied.  If the goal is to construct a complete sample,
then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
resolved binary companion \citep{Bouma_2022a}; the latter is on a
grazing orbit \citep{2014ApJ...783...53M}.  We leave selection for or
against such cases as a decision to the user.

@@

{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3
renormalized unit weight error (RUWE) as a function of other stellar
parameters, and flag stars with RUWE$>$1.4 as possible binaries.  Such
astrometric outliers can be either bona fide astrometric binaries, or
more often are marginally resolved point sources for which a
single-source PSF model provides a poor fit.

@@

{\it Candidate pulsators and close-in binaries (bit
10)}---\citeauthor{Santos_2021} included in a flag for candidate
``classical pulsators'' (e.g.\ Cepheids) and close-in binaries.
Visual inspection shows that although this flag selects many bona fide
objects in these classes, it also selects most $P_{\rm
rot}$$\lesssim$5\,day bona fide young planets.  These objects are
neither classical pulsators nor close binaries.  While we propagated
this flag into our set of quality flags, we therefore generally ignore
it.

@@

@@

@@

@@

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates
of main sequence stars.  Binarity is one example: even
intermediate-separation ($\sim$10\,AU) binaries are biased toward
rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
were interested in not only the overall age distribution of the Kepler
field, but also in the reliability of individual ages for young planet
hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
provides one path toward this end.  The other method is to consider
the region in which we do have overlapping sensitivity between lithium
and rotation -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
for such stars include \ltonegyrhighqconfirmedtwosided\ planets
orbiting ``high-quality'' stars for which the two age indicators are
both constraining and in agreement (with $Q_{\rm star}$ bits zero
through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a precise
lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

In this section we discuss the systems with confirmed planets that
have nominally discrepant rotation and lithium-based ages.  We find
that only one of them, Kepler-786, is interesting.

@@

@@
